---
title: "Analysis on Health Data for Bellabeat"
author: "Nadia Shureih"
date: "2024-07-16"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

### About Bellabeat

Bellabeat is a self described "pioneer in the fem-tech realm." Focused
on finding an intersection between women's wellness and women's fashion,
the company works to aid women in tracking a myriad of personal health
data. Although a small company, founder Urška Sršen believes that
analyzing health fitness data could unlock new areas of growth for the
company.

### Business Task

Analyze trends in health tracking smart device usage and provide
recommendations and opportunity for growth.

### Downloading Packages

#### Importing Libraries

```{r, echo=FALSE}
library("here")
library("skimr")
library("janitor")
library ("dplyr")
library("tidyverse")
library("ggplot2")
library("kableExtra")
library("knitr")
```

### Cleaning Data

#### Taking a Look at the Data

```{r}
sleep <- read.csv('sleepDay_merged.csv')
weight <- read.csv('weightLogInfo_merged.csv')
daily_activity <-read.csv('dailyActivity_merged.csv')
```

```{r}
head(sleep)
head(weight)
head(daily_activity)
```

#### Cleaning Duplicates

```{r}
sum(duplicated(daily_activity))
sum(duplicated(sleep))
sum(duplicated(weight))
```

```{r}
daily_activity <- daily_activity %>%
  distinct() %>%
  drop_na()

sleep <- sleep %>%
  distinct() %>%
  drop_na()
```

```{r}
sum(duplicated(sleep))
```

#### Creating Cohesive Date Formatting

```{r}
daily_activity <- daily_activity %>% 
  rename(Date = ActivityDate) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

sleep <- sleep %>% 
  rename(Date = SleepDay) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

weight <- weight %>% 
  select(-LogId) %>% 
  mutate(Date=as.Date(Date, format = "%m/%d/%y")) %>% 
  mutate(IsManualReport = as.factor(IsManualReport))
```

#### Cleaning Column Names

```{r, results="hide"}
clean_names(daily_activity)
clean_names(sleep)
clean_names(weight)
```

```{r}
daily_activity <- rename_with(daily_activity, tolower)
sleep <- rename_with(sleep, tolower)
weight <- rename_with(weight, tolower)
```
### Summarizing the Datasets

#### How many participants are in each dataset?

I wanted to get an understanding of how big each dataset is before
developing my analysis further and focusing in on the specifics.

```{r}
n_distinct(daily_activity$id)
n_distinct(sleep$id)
n_distinct(weight$id)
```

There are 35 participants in the activity data, 24 in the sleep data,
and 8 in weight data.

#### Getting a Summary of the Datasets

##### Daily Activity Summaries

```{r}
daily_activity %>% 
  select(totalsteps, veryactiveminutes, veryactivedistance, calories) %>% 
  summary()
```

```{r}
daily_activity %>% 
  select(sedentaryminutes, sedentaryactivedistance, fairlyactiveminutes, moderatelyactivedistance, lightlyactiveminutes, lightactivedistance) %>% 
  summary()
```

##### Some Conclusions from the Above Data

The average steps is 6547 (compare with the recommended 10,000 steps per
day). The average calories burned is 2189 (though this is wholy
dependent on age and gender).

Average amount of very active minutes is about 16.5, and very active
distance is about 1.18m. Average amount of sedentary minutes is about
995.3 Average amount of fairly active minutes is about 13.07, and
moderately active distance is about 0.48m. Average amount of lightly
active minutes is about 170.1, and very active distance is about 2.89m.

People spend the most amount of time sedentary, and are able to complete
more light activity next. People who complete light active are able to
go longer distances For people who are active, they will probably spend
more time in 'very active' than 'fairly active' meaning maybe they are
working harder in those workouts.

#### Sleep Summaries

```{r}
sleep %>% 
  select(totalsleeprecords, totalminutesasleep, totaltimeinbed) %>% 
  summary()
```

The average time asleep is 419.2 minutes or about 6.9 hours the average
time in bed is 458.5 minutes or about 7.5 hours.

#### Weight Summaries

```{r}
weight %>% 
  select(weightpounds, fat, bmi) %>% 
  summary()
```

The average weight is about 158.8lbs, average bmi is 25.19 (overweight),
which is understandable as someone most likely to be tracking their
weight would be someone who is overweight and trying to lose weight.

#### Merging Data Sets

```{r}
merged_data <- merge(merge(daily_activity, sleep, by= c('id','date'), all= TRUE), weight, by = c('id', 'date'), all= TRUE)

merged_activity_sleep <- merge(daily_activity, sleep, by=c('id','date'))
```

#### Removing Extra Variables

```{r}
merged_data <- merged_data %>% 
  select(-c(trackerdistance, totalsleeprecords, weightkg, ismanualreport))

merged_activity_sleep <- merged_activity_sleep %>% 
  select(-c(trackerdistance, totalsleeprecords))
```

Creating a data set which shows the averages of the users steps,calories, and sleep.

```{r}
daily_average <- merged_activity_sleep %>% 
  group_by(id) %>% 
  summarise(average_daily_steps = mean(totalsteps), average_daily_calories = mean(calories), average_daily_sleep = mean(totalminutesasleep))

average_steps <- daily_activity %>% 
  group_by(id)
```

### Visualizations

#### Understanding the User Data Base

```{r, echo= FALSE}
user_type <- daily_average %>% 
  mutate(user_type = case_when(
    average_daily_steps < 5000 ~ "sedentary",
    average_daily_steps >= 5000 & average_daily_steps < 7499 ~ "lightly active",
    average_daily_steps >= 7500 & average_daily_steps < 9999 ~ "fairly active", 
    average_daily_steps >= 1000 ~ "very active"
  ))

user_type_percent <- user_type %>% 
  group_by(user_type) %>% 
  summarise(total = n()) %>% 
  mutate(totals = sum(total)) %>% 
  group_by(user_type) %>% 
  summarise(total_percent = total/totals) %>% 
  mutate(labels = scales:: percent(total_percent))

user_type_percent$user_type <- factor(user_type_percent$user_type, levels = c("very active", "fairly active","lightly active", "sedentary"))
```

This pie chart then shows the different types of users based on activity:

```{r, echo=FALSE}
user_type_percent %>%
  ggplot(aes(x="",y=total_percent, fill=user_type)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  scale_fill_manual(values = c("#85e085","#e6e600", "#ffd480", "#ff8080")) +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))+
  labs(title="User type distribution")

```

Here we can see that a majority of users are actually within the fairly
active category, with roughly equal parts of the other types of
activity. **Bellabeat should aim to market towards people who are
already quite active, as well as market towards people inclining towards
improving their activity levels.**

```{r, echo=FALSE}
sleep_type <- daily_average %>% 
  mutate(sleep_type = case_when(
    average_daily_sleep < 360 ~ "poor",
    average_daily_sleep >= 360 & average_daily_sleep < 420 ~ "average",
    average_daily_sleep > 420 ~ "good"
  ))

sleep_type_percent <- sleep_type %>% 
  group_by(sleep_type) %>% 
  summarise(total = n()) %>% 
  mutate(totals = sum(total)) %>% 
  group_by(sleep_type) %>% 
  summarise(total_percent = total/totals) %>% 
  mutate(labels = scales:: percent(total_percent))

sleep_type_percent$sleep_type <- factor(sleep_type_percent$sleep_type, levels = c("poor", "average","good"))

sleep_user_type <- merge(sleep_type, user_type, by = "id") %>% 
  select(-c("average_daily_steps.y", "average_daily_calories.y", "average_daily_sleep.y"))
```

The following pie chart differentiates users based on their sleep type.

```{r, echo=FALSE}
sleep_type_percent %>%
  ggplot(aes(x="",y=total_percent, fill=sleep_type)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  scale_fill_manual(values = c("red","green", "pink")) +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))+
  labs(title="Sleep type distribution")
```

From here we see that the average user gets good sleep, although a large
portion of users get poor sleep. **Bellabeat can encourage better sleep
health for its users, perhaps by notifying users when it is time to
'wind down' or by giving users sleep tips to build better sleep
habits.**

The following bar graph shows the proportions of sleep types in the
different user types.

```{r, echo=FALSE}
ggplot(sleep_user_type, aes(user_type, fill=sleep_type)) +
  geom_bar() +
labs(title="Proportion of Sleep Types amongst User Types", x= "user type", y= "proportion of sleep types")

```

Here we see that fairly active users make up a majority of people with
'good' sleep, whereas poor sleep is relatively consistent amongst
activity levels. Perhaps ***encouraging users to increase their activity
levels will help their sleep trends***, although other factors may be at
play such as time of activity, caffeine intake, caloric intake, etc.
***Bellabeat should consider taking holistic data of their users to provide better
individualized health tips.***

#### Activity during the week

Total Steps by day of the week

```{r, echo=FALSE}
merged_data %>% 
  mutate(weekdays = weekdays(date)) %>% 
  select(weekdays, totalsteps) %>% 
  mutate(weekdays = factor(weekdays, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))) %>% 
  drop_na() %>% 
  ggplot(aes(weekdays, totalsteps, fill = weekdays)) +
  geom_boxplot()+
  labs(title="Total Steps by Day of the Week", y= "total steps")
```

This box plot shows us the amount of activity one might get during the
week, and specifically tells us that people in this data set are
consistent throughout the week but more sedentary on Sundays.
**Bellabeat can encourage user activity on these days, and encourage
rest days on Sundays to provide its users with moderation and time to
recuperate.**

Calories burned by day of the week.

```{r, echo=FALSE}
merged_data %>% 
  mutate(weekdays = weekdays(date)) %>% 
  select(weekdays, calories) %>% 
  mutate(weekdays = factor(weekdays, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))) %>% 
  drop_na() %>% 
  ggplot(aes(weekdays, calories, fill = weekdays)) +
  geom_boxplot() +
  labs(title="Calories Burned by Day of the Week")
```

Again, this box plot reiterates that people are less inclined to workout
on a Sunday, and more-so throughout the week.

#### Sleep

Sleep vs. Sedentary Minutes

```{r, echo=FALSE}
ggplot(data=merged_activity_sleep, aes(x=totalminutesasleep, y=sedentaryminutes)) + geom_point() + geom_smooth()+
  labs(title="Sedentary Minutes vs. Total Minutes Asleep", x= "Total Minutes Asleep", y= "Sedentary Minutes")
```

The above graph shows us that more sedentary minutes may roughly
correlate to less sleep. Taking our prior analysis from the user type
distribution graphs, **Bellabeat should consider promoting and
encouraging users with poor sleep to walk more or attain more daily
physical activity.**

lets compare with other activity types:

```{r, echo=FALSE}
merged_activity_sleep %>% 
  filter(veryactiveminutes != 0, fairlyactiveminutes != 0, lightlyactiveminutes != 0) %>%
  ggplot(aes(x=totalminutesasleep)) +
  geom_smooth(aes(y=veryactiveminutes), color="blue") +
  geom_smooth(aes(y=fairlyactiveminutes), color="red") +
  geom_smooth(aes(y=lightlyactiveminutes), color="green") +
  geom_smooth(aes(y=sedentaryminutes), color="purple")+
  labs(title="Activity Levels vs. Total Minutes Asleep", x = "Total Minutes Asleep", y = "Activity Levels")+
  annotate(geom="text",x=750, y=800, label="sedentary minutes")+
  annotate(geom="text",x=700, y=350, label="lightly active minutes")+
  annotate(geom="text",x=400, y=100, label="very active minutes")+
  annotate(geom="text",x=400, y=-5, label="fairly active minutes")
```

Here we see that sedentary minutes are truly the biggest factor in total
sleep.

### Summarizing Recommendations

A majority of fitness device users are already quite active. Bellabeat
can market to this demographic by creating devices which would benefit
those who are fairly active, **possibly by making lightweight fitness
devices that will individuals track their health data without getting in
the way of their workouts.** Bellabeat could also attempt to open up
their market to those who are not currently active, by **creating
advertisements which promote 'getting started' or 'jumping back into'
their health and fitness journey.**

Among this demographic, **Bellabeat might curate their device to make it
on the users schedule.** By assessing the above data on weekly steps and
caloric intake, **Bellabeat can encourage exercise throughout the week,
and encourage rest or cheat days on Weekends (particularly Sunday).** By
doing so, Bellabeat can market themselves as a self care app, **focused
on creating a schedule that works for the individual.**

Bellabeat can also work towards improving user's sleep trends **by
encouraging daily walking or activity through notifications**, letting
the user know how many steps they have left for the day or encouraging a
workout when their device notices them being sedentary for an extended
period of time.

**Bellabeat can also use their app to post health recommendations or
fitness blogs**, which can encourage more activity or better sleep
habits.
